---
title: "SOK-3022 Presentation"
subtitle: "CO2 VS GDP"
date: last-modified
date-format: "DD-MM-YYYY"
echo: false
warning: false
format: 
  revealjs:
    auto-stretch: false
    theme: moon
    smaller: true
---

## Introduction

**The relationship between CO2 and GDP**

We are going to explore the relationship between CO2 emissions and GDP using time series data. The analysis will be conducted using the Johansen cointegration test and Vector Error Correction Model (VECM) to understand both the long-run and short-run dynamics between these variables.

**Developing countries**

Further we are choosing to look at developing countries, to see if the relationship between CO2 emissions and GDP differs from that of developed countries. This is important as developing countries may have different economic structures and environmental policies that could influence this relationship.

**China and USA**

The two countries we will focus on are for non-OECD: China and the for OECD country we are choosing USA. China is the largest emitter of CO2 and has experienced rapid economic growth, while the USA has a long history of industrialization and is also a major emitter.

## Slide 1

You will receive a complete code file and sample data for replication. For your presentation, you'll only need to make minor adjustments to the code and select a few results to showcase. The primary goal of the presentation is to gather feedback on your preliminary research idea and choose one of the research topics covered in this module or the next.

You can prepare for the lecture in advance by reading (1) Chapters 9, 12, and 13 in Principle of Econometrics and (2) recommended literature in this Module.




                                                      
```{r}

rm(list=ls()) # Clear the environment
# Install/Load necessary libraries
library(readxl)   # Import xlsx data
library(tseries)  # For ADF test
library(car)      # Choose lag in VAR
library(urca)     # For Johansen test
library(vars)     # For VECM wo SE
library(tsDyn)    # For VECM with SE for Alpha
```
                                                      
```{r, output = FALSE}
D.CI <- read_excel("Lecture1/other/Data_World_Development_Indicators.xlsx", sheet='tilR');
D.CI = as.data.frame (D.CI); nrow(D.CI) # head(D.CI)
```
```{r, output = FALSE}                                                    
# wtf
D.CI1 = D.CI[D.CI[,4]%in%c('EN.GHG.CO2.MT.CE.AR5','NE.GDI.TOTL.KD','NY.GDP.MKTP.KD'), ]; nrow(D.CI)
D.CI1[,1]=D.CI1[,3]=NULL

varialbe = as.factor(D.CI1[,2]); levels(varialbe)
levels(varialbe) = c('CO2','Fin','GDP')                                                   
D.CI1[,2] = varialbe

names(D.CI1) = c('Country','Series',c(1960:2023))# str(D.CI1)
```
```{r}
D.CI2 = reshape(D.CI1, direction ='long', varying=list (names(D.CI1)[3:ncol(D.CI1)] ), v.names = "v", idvar = c("Country",'Series'), timevar='Year', time=1960:2023)
D.CI3 = reshape(D.CI2, idvar=c("Country","Year"), v.names=c('v'), timevar=c("Series"),direction="wide") ;  # head(D.CI3)
```

```{r, output = FALSE}
unique(D.CI3$Country)

Dat = D.CI3[D.CI3$Country=='CHN',]; nrow(Dat) # select a country
    sort(Dat$Year)
Dat = Dat

Dat = Dat[Dat$v.CO2>0, ]; nrow(Dat)
Dat = Dat[Dat$v.GDP>0, ]; nrow(Dat)
Dat = Dat[Dat$v.Fin>0, ]; nrow(Dat)

unique(Dat$Year)

Year = Dat$Year
```

Bla bla si noe her

```{r}
CO2 = log(Dat$v.CO2); summary(CO2)
FD = log(Dat$v.Fin); summary(FD)
GDP = log(Dat$v.GDP); summary(GDP)
```

## Slide 2

Bla bla si noe her

```{r}
adf.test(CO2)
adf.test(FD)
adf.test(GDP)
```
## Slide 3

Si noe her

```{r}
#$ Johansen: CO2 and FD  # ?ca.jo()
J_test1 <- ca.jo(data.frame(CO2,FD), type = "eigen", ecdet = c("const"), K = 2); summary(J_test1)
```
## Slide 4

Si noe her

```{r}
J_test2 <- ca.jo(data.frame(CO2,FD), type = "trace", ecdet = c("const"), K = 2); summary(J_test2)
```

```{r}
#$ Johansen: CO2 and GDP
J_test3 <- ca.jo(data.frame(CO2,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test3)
```

## Slide 5

Si noe her

```{r}
J_test4 <- ca.jo(data.frame(CO2,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test4)
```


```{r}
#$ Johansen: FD and GDP
J_test5 <- ca.jo(data.frame(FD,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test5)
```

## Slide 6

si noe her

```{r}
J_test6 <- ca.jo(data.frame(FD,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test6)
```


## Slide 7

si noe her

```{r}
VARselect(data.frame(CO2,FD,GDP), lag.max = 10, type = "trend")$selection

J_test7 <- ca.jo(data.frame(CO2,FD,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test7)
```

```{r}
J_test8 <- ca.jo(data.frame(CO2,FD,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test8)
```

## Slide 8

si noe her

```{r}
# VECM Function CO2 and FD

Vecm2 = VECM(data.frame(CO2,FD), lag = 1, r=1, estim = "ML",include=c('const') )# ?VECM()
summary(Vecm2)
```

## Slide 9

si noe her

```{r}
#$ long-run, beta

round(Vecm2$model.specific$beta,3)
cov = cov( residuals (Vecm2))
t_stats <- Vecm2$model.specific$beta / sqrt(diag(cov))
p_values <- 2 * (1 - pt(abs(t_stats), df = length(CO2) - 2))
p_values
```

## Slide 10

si noe her

```{r}
#$ short-run, alpha

round ( summary(Vecm2)$coefMat, 3)
```

```{r}
# VECM for CO2, FD, and GDP
Vecm2 = VECM(data.frame(CO2,FD,GDP), lag = 1, r=1, estim = "ML",include=c('const') )# ?VECM()
summary(Vecm2)
```

## Slide 11

si noe her

```{r}
# long-run, beta

round(Vecm2$model.specific$beta,3)
cov = cov( residuals (Vecm2))
t_stats <- Vecm2$model.specific$beta / sqrt(diag(cov))
p_values <- 2 * (1 - pt(abs(t_stats), df = length(CO2) - 2))
p_values
```

```{r}
# short-run, alpha
round ( summary(Vecm2)$coefMat, 3)
```

```{r}
Df = D.CI3[D.CI3$Country=='USA',]; nrow(Dat) # select a country
    sort(Df$Year)
Df = Df

Df = Df[Df$v.CO2>0, ]; nrow(Dat)
Df = Df[Df$v.GDP>0, ]; nrow(Dat)
Df = Df[Df$v.Fin>0, ]; nrow(Dat)

unique(Df$Year)

Year = Df$Year
```

Bla bla si noe her

```{r}
CO2 = log(Df$v.CO2); summary(CO2)
FD = log(Df$v.Fin); summary(FD)
GDP = log(Df$v.GDP); summary(GDP)
```

## Slide 2

Bla bla si noe her

```{r}
adf.test(CO2)
adf.test(FD)
adf.test(GDP)
```
## Slide 3

Si noe her

```{r}
#$ Johansen: CO2 and FD  # ?ca.jo()
J_test1 <- ca.jo(data.frame(CO2,FD), type = "eigen", ecdet = c("const"), K = 2); summary(J_test1)
```
## Slide 4

Si noe her

```{r}
J_test2 <- ca.jo(data.frame(CO2,FD), type = "trace", ecdet = c("const"), K = 2); summary(J_test2)
```

```{r}
#$ Johansen: CO2 and GDP
J_test3 <- ca.jo(data.frame(CO2,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test3)
```

## Slide 5

Si noe her

```{r}
J_test4 <- ca.jo(data.frame(CO2,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test4)
```


```{r}
#$ Johansen: FD and GDP
J_test5 <- ca.jo(data.frame(FD,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test5)
```

## Slide 6

si noe her

```{r}
J_test6 <- ca.jo(data.frame(FD,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test6)
```


## Slide 7

si noe her

```{r}
VARselect(data.frame(CO2,FD,GDP), lag.max = 10, type = "trend")$selection

J_test7 <- ca.jo(data.frame(CO2,FD,GDP), type = "eigen", ecdet = c("const"), K = 2); summary(J_test7)
```

```{r}
J_test8 <- ca.jo(data.frame(CO2,FD,GDP), type = "trace", ecdet = c("const"), K = 2); summary(J_test8)
```

## Slide 8

si noe her

```{r}
# VECM Function CO2 and FD

Vecm2 = VECM(data.frame(CO2,FD), lag = 1, r=1, estim = "ML",include=c('const') )# ?VECM()
summary(Vecm2)
```

## Slide 9

si noe her

```{r}
#$ long-run, beta

round(Vecm2$model.specific$beta,3)
cov = cov( residuals (Vecm2))
t_stats <- Vecm2$model.specific$beta / sqrt(diag(cov))
p_values <- 2 * (1 - pt(abs(t_stats), df = length(CO2) - 2))
p_values
```

## Slide 10

si noe her

```{r}
#$ short-run, alpha

round ( summary(Vecm2)$coefMat, 3)
```

```{r}
# VECM for CO2, FD, and GDP
Vecm2 = VECM(data.frame(CO2,FD,GDP), lag = 1, r=1, estim = "ML",include=c('const') )# ?VECM()
summary(Vecm2)
```

## Slide 11

si noe her

```{r}
# long-run, beta

round(Vecm2$model.specific$beta,3)
cov = cov( residuals (Vecm2))
t_stats <- Vecm2$model.specific$beta / sqrt(diag(cov))
p_values <- 2 * (1 - pt(abs(t_stats), df = length(CO2) - 2))
p_values
```

```{r}
# short-run, alpha
round ( summary(Vecm2)$coefMat, 3)
```

## Ferdig

## Ferdig
